---
title: "PostQC.Normalization.Processing"
author: "Angie Boysen"
date: "October 18, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(ggplot2)
library(tidyr)
require(graphics); require(grDevices)
library(Hmisc)
library(gtools)
library(cowplot)
require(RColorBrewer)
library(xlsx)
library(readr)
library(plotly)
library(stringr)
library(GGally)
library(dplyr)
```

##Exploring the HILIC diatom samples

#Import data - set filename with QC output here!
```{r}
filename <- "QC_outputExampleSkylineOutput.csv"
mydata.overloaded <- read_csv(filename,comment = "#")
mydata <- mydata.overloaded[,2:9]
glimpse(mydata)
```

##Read in Internal Standard list
Change the master list if you need to!
```{r}
IS.cat <- read_csv("HILIC_MasterList_Example.csv")  
IS.cat <- IS.cat %>% filter(Group == "Internal Std") %>% 
     select(`Compound Name`) %>% 
     unique()
IS.list <- IS.cat[[1]]
IS.list
```

##look at extraction replication
if things don't look good with some IS make sure to fix them
or at least not include them as possibilities for normalization
```{r, echo=FALSE}
IS.dat <- mydata[grepl(paste(IS.list, collapse="|"), mydata$Compound.Name),]
IS.dat <- IS.dat %>% select(Compound.Name, Area, Replicate.Name)
ggplot(IS.dat, aes(x=Replicate.Name, y=Area)) + geom_bar(stat="identity") + facet_wrap( ~Compound.Name, scales="free_y")+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5), legend.position = "top")  + ggtitle("IS Raw Areas")
```

##Pull out compounds that were detected
```{r detected?, echo=F}
Compds.detect <- mydata %>%      
     filter(!grepl("Blk",Replicate.Name)) %>%
     select(Replicate.Name, Compound.Name, Area) %>%
     group_by(Compound.Name) %>%
     summarize(percent.present = sum(!is.na(Area))/n())
# write.csv(Compds.detect,"in.samples.csv")
```

##Rename Runs
Name structure must be:
Date_type_ID_replicate
```{r echo=FALSE}
mydata <- mydata %>% mutate(Replicate.Name = Replicate.Name %>%
                                str_replace("_PooledDiatoms",
                                             "PooledDiatoms") %>%
                                 str_replace("0801_FilterBlank",
                                             "0801FilterBlank_1") %>%
                                 str_replace("Blk_9",
                                             "Blk_Blk_9") %>%
                                 str_replace("Np", "Np_") %>%
                                 str_replace("NP", "Np_") %>%
                                 str_replace("Pt", "Pt_") %>%
                                 str_replace("Tp", "Tp_") %>%
                                 str_replace("TP", "Tp_") %>%
                                 str_replace("160802_Smp_Pt_2", 
                                             "160802_Smp_Pt_12"))
```

##Calculate mean values for each IS
```{r ISmeans}
IS.means <- IS.dat %>% filter(!grepl("Blk", Replicate.Name)) %>%
     group_by(Compound.Name) %>%
     summarize(ave = mean(Area, rm.na=T))
```

##Normalize to each internal Standard
Actually scaling/adjusting by dividing by IS peak area and multiplying by the average IS peak area.
Output is a dataframe 'area.norm' that has each peak adjusted
to each possible internal standards
```{r echo=FALSE}
wArea<- mydata %>% 
     select(Replicate.Name,Compound.Name,Area) %>%
     spread(key=Compound.Name, value=Area) %>%
     as.data.frame

this.IS <- IS.list[1]
area.norm <- wArea[,-1] %>% 
          sapply(FUN = function(x) x/wArea[,grep(this.IS,
                                                 names(wArea))]) %>%
          as_data_frame %>% mutate(Replicate.Name = wArea$Replicate.Name) %>%
          gather(Compound,Area_Norm, -Replicate.Name)
this.mean <- IS.means %>% filter(Compound.Name==this.IS) %>% 
     select(ave) %>% as.numeric
area.norm <- area.norm %>% mutate(Area_Norm = Area_Norm*this.mean)
key <- ncol(area.norm)
count <- length(which(!is.na(area.norm$Area_Norm)))/
               length(unique(area.norm$Replicate.Name))
names(area.norm)[key] <- paste(this.IS,"Norm.Area",sep=".")
# print(paste(1, this.IS, count, sep="-"))

for (i in 2:length(IS.list)){
     this.IS <- IS.list[i]
     if(length(wArea[,grep(this.IS, names(wArea))])!=0){
          this.norm <- wArea[,-1] %>% 
               sapply(FUN = function(x) x/wArea[,grep(this.IS,
                                                      names(wArea))]) %>%
               as_data_frame %>% 
               mutate(Replicate.Name = wArea$Replicate.Name) %>%
               gather(Compound,Area_Norm, -Replicate.Name)
          this.mean <- IS.means %>% filter(Compound.Name==this.IS) %>% 
               select(ave) %>% as.numeric
          this.norm <- this.norm %>% mutate(Area_Norm = Area_Norm*this.mean)
          key <- ncol(area.norm)
          area.norm[,key+1] <- this.norm$Area_Norm
          names(area.norm)[key+1] <- paste(this.IS,"Norm.Area",sep=".")
          count <- length(which(!is.na(this.norm$Area_Norm)))/
               length(unique(this.norm$Replicate.Name))
          # print(paste(i, this.IS, count, sep="-"))
     }
}
glimpse(area.norm)
```

#Break Up the Names
Name structure must be:
Date_type_ID_replicate
```{r echo=FALSE}
mydata_new <- area.norm %>% separate(Replicate.Name, 
                                      c("runDate",
                                        "type","SampID","replicate"),"_") %>%
     mutate(Run.Cmpd = paste(area.norm$Replicate.Name,area.norm$Compound))
mydata <- mydata %>% mutate(Run.Cmpd = paste(Replicate.Name, Compound.Name))

dat <- full_join(mydata, mydata_new)
dat <- dat %>% select(-Compound)
# glimpse(dat)
```

##Compare normalizations
get mean, sd, and rsd for each normalization of each compound for each sample type.
Output is a dataframe called 'rsd.stats'
```{r}
no.blank.dat <- dat %>% filter(type =="Smp" | type =="Poo") %>%
     filter(SampID!="B")

rsd.stats <- no.blank.dat %>% select(-Replicate.Name, -Retention.Time,
                                     -Notes, -Height, -IR, -S.N, -Run.Cmpd,
                                     -runDate, -replicate) %>%
     gather(Normer, Value, -Compound.Name, -type, - SampID) %>%
     group_by(Compound.Name, type, SampID, Normer) %>%
     summarize(m = mean(Value, na.rm=T), 
                sd = sd(Value, na.rm=T), rsd = sd/m)
glimpse(rsd.stats)
```

##Cleanup RSD data and add a minimizing column 
One based on the multiple injetions of the pooled sample "PooModel"
and one based on the injections of the biological replicates of samples "SmpModel"
```{r newModeMaker}
rsd.clean <- rsd.stats %>% filter(!is.na(m)) %>%
     filter(Normer!="Area")
SmpModel <- rsd.clean %>% filter(type=="Smp") %>%
     select(-m, -sd) %>%
     group_by(Compound.Name, Normer) %>%
     summarize(Mean.rsd = mean(rsd, na.rm=T)) %>%
     summarize(SmpModelRSD = min(Mean.rsd),
               Smp.Picked.IS = unique(Normer)[which.min(Mean.rsd)])
PooModel <- rsd.clean %>% filter(type=="Poo") %>%
     select(-m, -sd) %>%
     group_by(Compound.Name, Normer) %>%
     summarize(Mean.rsd = mean(rsd, na.rm=T)) %>%
     summarize(PooModelRSD = min(Mean.rsd),
               Poo.Picked.IS = unique(Normer)[which.min(Mean.rsd)])
Models <- full_join(PooModel, SmpModel)
glimpse(Models)
```

##Merge Models and RSD data all together
```{r}
rsd.total <- full_join(rsd.stats, Models)
```

##Adjust the pooled model --> PooPlus model aka B-MIS model
so that it normalizes metabolites that we have internal standards
for to the appropriate internal standard. Otherwise keeps the pooled
model. 
Otherwise uses the PooModel for all compounds detected in the Pooled samples.
If a compound was detected in the sample but not the pooled then use the SmpModel.
Names this new complied model "PooPlus" (aka B-MIS)
Output of this section is 'rsd.total': A dataframe with mean, sd, and rsd for each compound, sample type, and normalization combination and a column indicating the poo, smp and pooPlus (B-MIS) internal standard selections 
```{r echo=FALSE}
rsd.total <- rsd.total %>%
     mutate(PooPlus.IS = Poo.Picked.IS) %>%
     mutate(PooPlusModelRSD = PooModelRSD)

for (i in 1:nrow(rsd.total)){
     cmpd <- rsd.total$Compound.Name[i]
     if(length(grep(cmpd, IS.list))>0){
          newIS <- paste0(IS.list[grep(cmpd, IS.list)],".Norm.Area")
          rsd.total$PooPlus.IS[i] <- newIS
          rsd.total$PooPlusModelRSD[i] <- filter(rsd.total, 
                                                 Compound.Name == cmpd, 
                                                 Normer == newIS, 
                                                 type == "Poo")$rsd
     }
     ## if the poo didn't pick a normalizing IS, use the samp picked one
     if(is.na(rsd.total$Poo.Picked.IS[i]) &
        !is.na(rsd.total$Smp.Picked.IS[i])){
          rsd.total$PooPlus.IS[i] <- rsd.total$Smp.Picked.IS[i]
          rsd.total$PooPlusModelRSD[i] <-
               rsd.total$SmpModelRSD[i]
     }
}
rsd.total <- rsd.total %>% 
     select(-PooModelRSD,-PooPlusModelRSD,-SmpModelRSD)
glimpse(rsd.total)
```

##Write out RSD data, if you want
```{r, eval=FALSE}
write.csv(rsd.total, "All_normalization_rsd_dat.csv")
```

##Get Actual Normalized data not just the rsds
```{r, echo=FALSE}
models <- rsd.total %>% ungroup %>%
     select(Compound.Name,
            Poo.Picked.IS,Smp.Picked.IS,PooPlus.IS) %>%
     group_by(Compound.Name) %>%
     summarize(PooModel.IS = unique(Poo.Picked.IS),
               SmpModel.IS = unique(Smp.Picked.IS),
               PooPlusModel.IS = unique(PooPlus.IS))
dat <- dat %>% filter(!is.na(Replicate.Name))
dat.join <- full_join(dat, models)

PooModel <- c()
SmpModel <-c()
PooPlusModel <- c()
for (i in 1:nrow(dat.join)){
     if(!is.na(dat.join$PooModel.IS[i])){
          PooModel <- c(PooModel,dat.join[i,dat.join$PooModel.IS[i]])
     } else {
          PooModel <- c(PooModel,NA)
     }
     if(!is.na(dat.join$SmpModel.IS[i])){
          SmpModel <- c(SmpModel,dat.join[i,
                                            dat.join$SmpModel.IS[i]])
     } else {
          SmpModel <- c(SmpModel,NA)
     }
     if(!is.na(dat.join$PooPlusModel.IS[i])){
          PooPlusModel <- c(PooPlusModel,dat.join[i,
                                            dat.join$PooPlusModel.IS[i]])
     } else {
          PooPlusModel <- c(PooPlusModel,NA)
     }
}
dat.join$PooModel <- PooModel
dat.join$SmpModel <- SmpModel
dat.join$PooPlusModel <- PooPlusModel
dat.join <- dat.join %>% unnest(PooModel) %>% unnest(SmpModel) %>%
     unnest(PooPlusModel)
glimpse(dat.join)
```

##Write out that full normalized data
The output of the B-MIS normalized data is in the column named PooPlusModel.
```{r, eval=FALSE}
new.filename <- paste("Normalized_data",filename,sep="_")
write.csv(dat.join, new.filename)
```
