---
title: "Best Matched Internal Standard for Targeted TQS Data"
author: "Boysen and Heal et al"
date: "July 31, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'pdf')


library(ggplot2)
library(tidyr)
require(graphics); require(grDevices)
library(Hmisc)
library(gtools)
library(cowplot)
require(RColorBrewer)
library(xlsx)
library(readr)
library(plotly)
library(stringr)
library(GGally)
library(dplyr)
```


#Import data - set filenames within this chunk for QC output, master list, and sample key here
```{r, error = FALSE, warning=FALSE, message=FALSE}
filename <- "QC_outputExampleSkylineOutput.csv"
mydata.overloaded <- read_csv(filename,comment = "#")
mydata <- mydata.overloaded %>% select(Replicate.Name:S.N) %>% 
  filter(!Notes %in% c("Standards"))
glimpse(mydata)

sample.key <- read.csv("sample.key.targeted.csv")
glimpse(sample.key)

MasterList <- "HILIC_MasterList_Example.csv"

FlagforIS <- "Internal Std" #Flag on your master list for the internal standards
```

#Internal Standard list that BMIS will match to
```{r, error = FALSE, warning=FALSE, message=FALSE}
IS.cat <- read_csv(MasterList) %>% filter(Group == FlagforIS) %>% 
     select(`Compound Name`) %>% 
     unique()
IS.list <- IS.cat[[1]]
print(IS.list)
```

#Look at extraction replication of the Internal Standards
If things don't look good with some IS or certain samples, make sure to fix them or at least not include them as possibilities for normalization
```{r,  error = FALSE, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
IS.dat <- mydata[grepl(paste(IS.list, collapse="|"), mydata$Compound.Name),]
IS.dat <- IS.dat %>% select(Compound.Name, Area, Replicate.Name)
ggplot(IS.dat, aes(x=Replicate.Name, y=Area)) + 
  geom_bar(stat="identity") + 
  facet_wrap( ~Compound.Name, scales="free_y")+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5, size = 5),
        axis.text.y = element_text(size = 7),
        legend.position = "top",
        strip.text = element_text(size = 7)) +
  ggtitle("IS Raw Areas")
```

#Generate a list of compounds that were detected
```{r, error = FALSE, warning=FALSE, message=FALSE}
Compds.detect <- mydata %>%      
     filter(!grepl("Blk",Replicate.Name)) %>%
     select(Replicate.Name, Compound.Name, Area) %>%
     group_by(Compound.Name) %>%
     summarise(percent.present = sum(!is.na(Area))/n())
print(Compds.detect$Compound.Name)
# write.csv(Compds.detect,"in.samples.csv")
```

#Rename Runs - optional chunk to adjust the naming of individual runs 
##EDIT This if your names do not follow the exact structure
Name structure must be: Date_type_ID_replicate
```{r, error = FALSE, warning=FALSE, message=FALSE}
# mydata <- mydata %>% mutate(Replicate.Name = Replicate.Name %>%
#                                 str_replace("_PooledDiatoms",
#                                              "PooledDiatoms") %>%
#                                  str_replace("0801_FilterBlank",
#                                              "0801FilterBlank_1") %>%
#                                  str_replace("Blk_9",
#                                              "Blk_Blk_9") %>%
#                                  str_replace("Np", "Np_") %>%
#                                  str_replace("NP", "Np_") %>%
#                                  str_replace("Pt", "Pt_") %>%
#                                  str_replace("Tp", "Tp_") %>%
#                                  str_replace("TP", "Tp_") %>%
#                                  str_replace("160802_Smp_Pt_2", 
#                                              "160802_Smp_Pt_12"))
print(unique(mydata$Replicate.Name))
```

#Calculate mean values for each IS
```{r ISmeans, error = FALSE, warning=FALSE, message=FALSE}
IS.means <- IS.dat %>% filter(!grepl("Blk", Replicate.Name)) %>%
     group_by(Compound.Name) %>%
     summarise(ave = mean(Area, rm.na=T))
IS.means <- rbind(IS.means, c("inj.vol",1)) %>%
     mutate(ave = as.numeric(ave))
glimpse(IS.means)
```

#Normalize to each internal Standard
Adjust peak areas to each internal standard.  Output is a dataframe 'area.norm' that has each peak adjusted to each possible internal standards
```{r, error = FALSE, warning=FALSE, message=FALSE}
wArea<- mydata %>% 
     select(Replicate.Name,Compound.Name,Area) %>%
     spread(key=Compound.Name, value=Area) %>% 
     as.data.frame
wArea.2 <- full_join(wArea, sample.key[,c("Sample.Name","Sample.Group","Bio.Normalization")], by = c("Replicate.Name"="Sample.Name")) %>%
     mutate(inj.vol = ifelse(Sample.Group=="Poo",Bio.Normalization,NA)) %>%
     select(-Sample.Group, -Bio.Normalization)

this.IS <- "inj.vol"
area.norm <- wArea.2[,-1] %>% 
          sapply(FUN = function(x) x/wArea.2[,grep(this.IS,
                                                 names(wArea.2))]) %>%
          as_data_frame %>% mutate(Replicate.Name = wArea.2$Replicate.Name) %>%
          gather(Compound,Area_Norm, -Replicate.Name)
this.mean <- IS.means %>% filter(Compound.Name==this.IS) %>% 
     select(ave) %>% as.numeric
area.norm <- area.norm %>% mutate(Area_Norm = Area_Norm*this.mean)
key <- ncol(area.norm)
count <- length(which(!is.na(area.norm$Area_Norm)))/
               length(unique(area.norm$Replicate.Name))
names(area.norm)[key] <- paste(this.IS,"Norm.Area",sep=".")

for (i in 1:length(IS.list)){
     this.IS <- IS.list[i]
     if(length(wArea.2[,grep(this.IS, names(wArea.2))])!=0){
          this.norm <- wArea.2[,-1] %>% 
               sapply(FUN = function(x) x/wArea.2[,grep(this.IS,
                                                      names(wArea.2))]) %>%
               as_data_frame %>% 
               mutate(Replicate.Name = wArea.2$Replicate.Name) %>%
               gather(Compound,Area_Norm, -Replicate.Name)
          this.mean <- IS.means %>% filter(Compound.Name==this.IS) %>% 
               select(ave) %>% as.numeric
          this.norm <- this.norm %>% mutate(Area_Norm = Area_Norm*this.mean)
          key <- ncol(area.norm)
          area.norm[,key+1] <- this.norm$Area_Norm
          names(area.norm)[key+1] <- paste(this.IS,"Norm.Area",sep=".")
          count <- length(which(!is.na(this.norm$Area_Norm)))/
               length(unique(this.norm$Replicate.Name))
          # print(paste(i, this.IS, count, sep="-"))
     }
}
glimpse(area.norm)
```

#Break Up the Names
Name structure must be:
Date_type_ID_replicate
```{r, error = FALSE, warning=FALSE, message=FALSE}
mydata_new <- area.norm %>% separate(Replicate.Name, 
                                      c("runDate",
                                        "type","SampID","replicate"),"_") %>%
     mutate(Run.Cmpd = paste(area.norm$Replicate.Name,area.norm$Compound))
mydata <- mydata %>% mutate(Run.Cmpd = paste(Replicate.Name, Compound.Name))

dat <- full_join(mydata, mydata_new)
dat <- dat %>% select(-Compound)
print(dat %>% select(runDate:replicate) %>% unique)
```

#Compare normalizations
Get mean, sd, and rsd for each normalization of each compound for each sample type.
Output is a dataframe called 'rsd.stats'
```{r, error = FALSE, warning=FALSE, message=FALSE}
no.blank.dat <- dat %>% filter(type =="Smp" | type =="Poo") %>%
     filter(SampID!="B")

rsd.stats <- no.blank.dat %>% select(-Replicate.Name, -Retention.Time,
                                     -Notes, -Height, -IR, -S.N, -Run.Cmpd,
                                     -runDate, -replicate) %>%
     gather(Normer, Value, -Compound.Name, -type, - SampID) %>%
     group_by(Compound.Name, type, SampID, Normer) %>%
     summarise(m = mean(Value, na.rm=T), 
                sd = sd(Value, na.rm=T), rsd = sd/m)
glimpse(rsd.stats)
```

##Cleanup RSD data and add a minimizing column 
One based on the multiple injetions of the pooled sample "PooModel"
and one based on the injections of the biological replicates of samples "SmpModel"
```{r,  error = FALSE, warning=FALSE, message=FALSE}
rsd.clean <- rsd.stats %>% filter(!is.na(m)) %>%
     filter(Normer!="Area")
SmpModel <- rsd.clean %>% filter(type=="Smp") %>%
     select(-m, -sd) %>%
     group_by(Compound.Name, Normer) %>%
     summarise(Mean.rsd = mean(rsd, na.rm=T)) %>%
     summarise(SmpModelRSD = min(Mean.rsd),
               Smp.Picked.IS = unique(Normer)[which.min(Mean.rsd)])
PooModel <- rsd.clean %>% filter(type=="Poo") %>%
     select(-m, -sd) %>%
     group_by(Compound.Name, Normer) %>%
     summarise(Mean.rsd = mean(rsd, na.rm=T)) %>%
     summarise(PooModelRSD = min(Mean.rsd),
               Poo.Picked.IS = unique(Normer)[which.min(Mean.rsd)][1])
Models <- full_join(PooModel, SmpModel)
glimpse(Models)

rsd.total <- full_join(rsd.stats, Models) %>%
     filter(!is.na(Compound.Name))
```

#Adjust the pooled model --> PooPlus model aka B-MIS model
Force BMIS to normalizes metabolites that we have internal standards for to the appropriate internal standard. Otherwise keeps the pooled model. 
Otherwise uses the PooModel for all compounds detected in the Pooled samples.
If a compound was detected in the sample but not the pooled then use the SmpModel.
Names this new complied model "PooPlus" (aka B-MIS)
Output of this section is 'rsd.total': A dataframe with mean, sd, and rsd for each compound, sample type, and normalization combination and a column indicating the poo, smp and pooPlus (B-MIS) internal standard selections 
```{r error = FALSE, warning=FALSE, message=FALSE}
rsd.total <- rsd.total %>%
     mutate(PooPlus.IS = Poo.Picked.IS) %>%
     mutate(PooPlusModelRSD = PooModelRSD)

for (i in 1:nrow(rsd.total)){
     cmpd <- rsd.total$Compound.Name[i]
     if(length(grep(cmpd, IS.list))>0){
          newIS <- paste0(IS.list[grep(cmpd, IS.list)],".Norm.Area")
          rsd.total$PooPlus.IS[i] <- newIS
          
     }
     ## if the poo didn't pick a normalizing IS, use the samp picked one
     if(is.na(rsd.total$Poo.Picked.IS[i]) &
        !is.na(rsd.total$Smp.Picked.IS[i])){
          rsd.total$PooPlus.IS[i] <- rsd.total$Smp.Picked.IS[i]
          rsd.total$PooPlusModelRSD[i] <-
               rsd.total$SmpModelRSD[i]
     }
}






rsd.total <- rsd.total %>% 
     select(-PooModelRSD,-PooPlusModelRSD,-SmpModelRSD)
glimpse(rsd.total)
#write.csv(rsd.total, "New_All_normalization_rsd_dat.csv") #Uncomment if you'd like to write out the RSD data

```

#Get Adjusted areas after choosing a BMIS
```{r, error = FALSE, warning=FALSE, message=FALSE}
models <- rsd.total %>% ungroup %>%
     select(Compound.Name, PooPlus.IS) %>%
     group_by(Compound.Name) %>%
     summarise(PooPlusModel.IS = unique(PooPlus.IS))
dat <- dat %>% filter(!is.na(Replicate.Name))
dat.join <- as.data.frame(full_join(dat, models)) %>%
  mutate(PooPlusModel = NA)

split.on.IS <- as.factor(dat.join$PooPlusModel.IS)
split.dat.join <- split(dat.join, split.on.IS)
for (i in 1:length(split.dat.join)){
  col.key <-  which(names(split.dat.join[[i]])==names(split.dat.join)[i])
  split.dat.join[[i]]$PooPlusModel <- split.dat.join[[i]][,col.key]
}

unsplit.dat.join <- unsplit(split.dat.join, split.on.IS)
glimpse(unsplit.dat.join)

new.filename <- paste("New_Normalized_data",filename,sep="_")
write.csv(unsplit.dat.join, new.filename)
print(paste("Your BMIS results are saved as", new.filename))
print("The output of the B-MIS normalized data is in the column named PooPlusModel.")

```
