get---
title: "QE.XCMS.CyanoAq.TM2P.Diatoms.BMIS"
author: "Angie Boysen"
date: "October 30, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'pdf')

library(ggplot2)
library(tidyr)
require(graphics); require(grDevices)
library(Hmisc)
library(gtools)
library(cowplot)
require(RColorBrewer)
library(readr)
library(plotly)
library(stringr)
library(GGally)
library(dplyr)
```


#Import data - set filenames within this chunk for xcms output, sample key, and ISdata
```{r, error = FALSE, warning=FALSE, message=FALSE}
filename <- "xset.filtered.csv"
xcms.dat <- read_csv(filename)
xcms.names <- names(xcms.dat)

ISfilename1 <- "IS_ResultsNeg.csv"
ISfilename2 <- "IS_ResultsPos.csv"
ISdatfull <- read_csv(ISfilename1) %>% rbind(read_csv(ISfilename2))

SampKeyfilename <- "sample.key.untargeted.csv"
```

#Read in Internal Standard data, add in injec_volume data from Sample Key
```{r, error = FALSE, warning=FALSE, message=FALSE}
IS.dat <- ISdatfull %>%
     select(`Replicate Name`, `Precursor Ion Name`, Area) %>%
     mutate(MassFeature = `Precursor Ion Name`) %>%
  select(-`Precursor Ion Name`)

SampKey <- read_csv(SampKeyfilename) %>%
  filter(Sample.Name %in% IS.dat$`Replicate Name`) %>%
  select(Sample.Name, Injec_vol) %>%
  filter(!is.na(Injec_vol))%>%
  mutate(MassFeature = "Inj_vol",
         Area = Injec_vol,
         `Replicate Name` = Sample.Name) %>%
  select(`Replicate Name`, Area, MassFeature)

IS.dat <- rbind(IS.dat, SampKey)
glimpse(IS.dat)
```


#Look at extraction replication of the Internal Standards
If things don't look good with some IS or certain samples, make sure to fix them or at least not include them as possibilities for normalization
```{r,  error = FALSE, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
ggplot(IS.dat, aes(x=`Replicate Name`, y=Area)) + 
  geom_bar(stat="identity") + 
  facet_wrap( ~MassFeature, scales="free_y")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5, size = 5), 
        axis.text.y = element_text(size = 7),
        legend.position = "top",
        strip.text = element_text(size = 7))+
  ggtitle("IS Raw Areas")
```

#Edit IS data if necessary
Optional if there are certain IS that aren't trustworth
```{r ,  error = FALSE, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
IS.dat <- IS.dat %>%
  filter (MassFeature != "Heavy Acetyl CoA") %>%
  filter (MassFeature != "Heavy Succinic Acid")
```


#Get untargeted data into long format
```{r,  error = FALSE, warning=FALSE, message=FALSE}
IS.dat <- IS.dat %>% mutate(Replicate.Name = `Replicate Name` %>%
                                str_replace("-","."))  %>%
  select(Area, Replicate.Name, MassFeature)

replicates <- paste0("X",unique(IS.dat$Replicate.Name)) #Only use this if you have numbers at the beginning of runs

col.key <-  xcms.names %in% replicates

xcms.long <- cbind(xcms.dat[,col.key],MassFeature = xcms.dat$MassFeature) %>%
     gather(Replicate.Name, Area, -MassFeature)
glimpse(xcms.long)
```

#Rename Runs - optional chunk to adjust the naming of individual runs 
##EDIT This if your names do not follow the exact structure
Name structure must be: Date_type_ID_replicate
```{r ,  error = FALSE, warning=FALSE, message=FALSE}
xcms.long <- xcms.long %>%
  mutate(Replicate.Name = Replicate.Name %>%
            str_replace("_Blk_Blk_M","_Blk_M")%>%
            str_replace("ExtractsFull_","Extracts_Full")%>%
           str_replace("ExtractsHalf_","Extracts_Half"))

IS.dat <- IS.dat %>%
  mutate(Replicate.Name = Replicate.Name %>%
           str_replace("17","X17") %>%
            str_replace("_Blk_Blk_M","_Blk_M")%>%
           str_replace("ExtractsFull_","Extracts_Full")%>%
           str_replace("ExtractsHalf_","Extracts_Half"))

print(unique(xcms.long$Replicate.Name))
```

#Calculate mean values for each IS
```{r ISmeans,  error = FALSE, warning=FALSE, message=FALSE}
IS.means <- IS.dat %>% filter(!grepl("_Blk_", Replicate.Name)) %>%
  mutate(MassFeature = as.factor(MassFeature))%>%
     group_by(MassFeature) %>%
     summarise(ave = mean(Area))
head(IS.means)
```

#Normalize to each internal Standard
Actually scaling/adjusting by dividing by IS peak area and multiplying by the average IS peak area.
Output is a dataframe 'area.norm' that has each peak adjusted
to each possible internal standards
```{r, error = FALSE, warning=FALSE, message=FALSE}
binded <- rbind(IS.dat, xcms.long)
wArea<- binded %>% 
     select(Replicate.Name, MassFeature, Area) %>%
     spread(key=MassFeature, value=Area) %>%
     as.data.frame

IS.list <- unique(IS.dat$MassFeature)
this.IS <- IS.list[1]
area.norm <- wArea[,-1] %>% 
          sapply(FUN = function(x) x/wArea[,grep(this.IS,
                                                 names(wArea))]) %>%
          as_data_frame %>% mutate(Replicate.Name = wArea$Replicate.Name) %>%
          gather(MassFeature,Area_Norm, -Replicate.Name)
this.mean <- IS.means %>% filter(MassFeature==this.IS) %>% 
     select(ave) %>% as.numeric
area.norm <- area.norm %>% mutate(Area_Norm = Area_Norm*this.mean)
key <- ncol(area.norm)
count <- length(which(!is.na(area.norm$Area_Norm)))/
               length(unique(area.norm$Replicate.Name))
names(area.norm)[key] <- paste(this.IS,"Norm.Area",sep=".")
# print(paste(1, this.IS, count, sep="-"))

for (i in 2:length(IS.list)){
     this.IS <- IS.list[i]
     if(length(wArea[,grep(this.IS, names(wArea))])!=0){
          this.norm <- wArea[,-1] %>% 
               sapply(FUN = function(x) x/wArea[,grep(this.IS,
                                                      names(wArea))]) %>%
               as_data_frame %>% 
               mutate(Replicate.Name = wArea$Replicate.Name) %>%
               gather(MassFeature,Area_Norm, -Replicate.Name)
          this.mean <- IS.means %>% filter(MassFeature==this.IS) %>% 
               select(ave) %>% as.numeric
          this.norm <- this.norm %>% mutate(Area_Norm = Area_Norm*this.mean)
          key <- ncol(area.norm)
          area.norm[,key+1] <- this.norm$Area_Norm
          names(area.norm)[key+1] <- paste(this.IS,"Norm.Area",sep=".")
          count <- length(which(!is.na(this.norm$Area_Norm)))/
               length(unique(this.norm$Replicate.Name))
          # print(paste(i, this.IS, count, sep="-"))
     }
}
glimpse(area.norm)
```


#Break Up the Names
Name structure must be:
Date_type_ID_replicate
```{r , error = FALSE, warning=FALSE, message=FALSE}
mydata_new <- area.norm %>% separate(Replicate.Name, 
                                      c("runDate",
                                        "type","SampID","replicate"),"_") %>%
     mutate(Run.Cmpd = paste(area.norm$Replicate.Name,area.norm$MassFeature))
binded <- binded %>% mutate(Run.Cmpd = paste(Replicate.Name, MassFeature))

dat <- full_join(binded, mydata_new)
print(dat %>% select(runDate:replicate) %>% unique)
# write.csv(dat, "IS.data.csv")
```

#Make a list of the internal standards, - the MIS, and RSD of pooled, RSD of smp for that
```{r}
cut.off <- 0.4

dat2 <- dat %>%
  filter(MassFeature %in% IS.dat$MassFeature) %>%
  select(-(Area:MassFeature))  %>%
  select(-(runDate:replicate)) %>%
  gather(key = "MIS", value = "Adjusted_Area", factor_key = TRUE, -Run.Cmpd) %>%
  left_join(dat %>% select(type, MassFeature, Run.Cmpd)) %>%
  mutate(Adjusted_Area = as.numeric(Adjusted_Area))

smpdat <- dat2 %>%
  filter(type == "Smp")%>%
  group_by(MassFeature, MIS) %>%
  summarise(RSD_ofSmp = sd(Adjusted_Area)/mean(Adjusted_Area))

alldat <- dat2 %>%
  filter(type == "Poo")%>%
  group_by(MassFeature, MIS) %>%
  summarise(RSD_ofPoo = sd(Adjusted_Area)/mean(Adjusted_Area)) %>%
  left_join(smpdat)

injectONlY <- alldat %>%
  filter(MIS == "Inj_vol.Norm.Area" ) %>%
  mutate(Orig_RSD = RSD_ofPoo)%>%
  select(-RSD_ofPoo, -RSD_ofSmp, -MIS)

injectONlY_toPlot <- alldat %>%
  filter(MIS == "Inj_vol.Norm.Area" ) 

newalldat <- left_join(alldat, injectONlY) %>%
  mutate(del_RSD = ( Orig_RSD - RSD_ofPoo),
         percRSD = del_RSD/Orig_RSD) %>%
  mutate(accept_MIS = (percRSD > cut.off))

write.csv(newalldat,  "IS_testResults.csv")



g <- ggplot()+
  geom_point(dat = newalldat, aes(x = RSD_ofPoo, y = RSD_ofSmp, colour = accept_MIS))+
  geom_point(dat = injectONlY_toPlot, aes(x = RSD_ofPoo, y = RSD_ofSmp), size = 2) +
  facet_wrap(~MassFeature)

g

ggsave("IS_experiment.pdf", plot = g, device = "pdf", width = 10, height = 10, units = "in")
```


#Make a summary of which internal standards were conidsterd "Acceptable" and "Best" Matching Internal Standards
```{r}
## This dataset is all MFs normalized to all ISs
# dat <- read_csv("IS_data.csv")%>%
#   select(-X1)

#Makes data Long
longdat <- dat %>%
  select(-(Area:MassFeature))  %>%
  select(-(runDate:replicate)) %>%
  gather(key = "MIS", value = "Adjusted_Area", factor_key = TRUE, -Run.Cmpd) %>%
  left_join(dat %>% select(type, MassFeature, Run.Cmpd)) %>%
  mutate(Adjusted_Area = as.numeric(Adjusted_Area))

#Look only the Pooled samples, to get a lowest RSD of the pooled possible (RSD_ofPoo), then choose which IS reduces the RSD the most (Poo.Picked.IS)
poodat <- longdat %>%
  filter(type == "Poo")%>%
  group_by(MassFeature, MIS) %>%
  summarise(RSD_ofPoo =  sd(Adjusted_Area)/mean(Adjusted_Area))
poodat <- poodat %>% left_join(poodat %>%
                                 group_by(MassFeature) %>%
                                 summarise(Poo.Picked.IS = unique(MIS)[which.min(RSD_ofPoo)][1]))

#Get the starting point of the RSD (Orig_RSD), calculate the change in the RSD, say if the MIS is acceptable
newpoodat <- left_join(poodat, poodat %>%
                         filter(MIS == "Inj_vol.Norm.Area" ) %>%
                         mutate(Orig_RSD = RSD_ofPoo) %>%
                         select(-RSD_ofPoo, -MIS)) %>%
  mutate(del_RSD = (Orig_RSD - RSD_ofPoo)) %>%
  mutate(percentChange = del_RSD/Orig_RSD) %>%
  mutate(accept_MIS = (percentChange > 0.4 & Orig_RSD > 0.1))  #HERE IS WHERE WE NEED TO DEFINE WHAT MAKES AND ACCEPTABLE MATCHED INTERNAL STANDARD

#Change the BMIS to "Inj_vol.Norm.Area" if the BMIS is not an acceptable - Adds a column that has the BMIS, not just Poo.picked.IS
NoAMIS_newpoodat_BMIS <- newpoodat %>%
  filter(MIS == Poo.Picked.IS) %>% #Pulling out just the pooplus is matches
  filter(accept_MIS == "FALSE") %>%
  mutate(FinalBMIS = "Inj_vol.Norm.Area") 

newpoodat_BMIS <- newpoodat %>%
              filter(MIS == Poo.Picked.IS) %>% 
              filter(accept_MIS == "TRUE") %>%
              mutate(FinalBMIS = Poo.Picked.IS) %>%
  bind_rows(NoAMIS_newpoodat_BMIS)

#Makes the full dataset from the fixed ones (where there is no AMIS)
FullDat_fixed <- left_join(newpoodat, newpoodat_BMIS %>%
                        select(MassFeature, FinalBMIS)) 
FullDat_fixed <- FullDat_fixed %>% left_join(FullDat_fixed %>%
              filter(FinalBMIS == MIS) %>%
              mutate(FinalRSD = RSD_ofPoo) %>%
              select(MassFeature, FinalRSD))

#Only get MFs with FinalRSD < 0.2
Good_MFs <- FullDat_fixed %>%
  filter(MIS == Poo.Picked.IS) %>%
  filter(FinalRSD < 0.2)

#Get number of mass features
MassFeatures <- length(Good_MFs$MassFeature)
print("Total mass features")
print(MassFeatures)

 #This will give us the number of mass features that were picked as a BMIS for each IS
BMIS_percent <- Good_MFs %>%
  filter(FinalBMIS != "Inj_vol.Norm.Area") %>%
  group_by(FinalBMIS) %>%
  summarise(MFs = n()) %>%
  mutate(PercentofBMIS = MFs/MassFeatures)  

BMIS_Summary2 <- Good_MFs %>%
  filter(FinalBMIS != "Inj_vol.Norm.Area")
MassFeatures2 <- length(BMIS_Summary2$MassFeature)
print("Mass features with BMIS")
print(MassFeatures2)


#This will give us the number of mass features that were picked as an AMIS for each IS
AMIS_percent <- FullDat_fixed %>%
  filter(MassFeature %in% Good_MFs$MassFeature) %>%
  filter(accept_MIS == "TRUE") %>%
  group_by(FinalBMIS) %>%
  summarise(MFs = n()) %>%
  mutate(PercentofAMIS = MFs/MassFeatures)  #This gives us the % of the MF chosen by each IS 

Summary <- left_join(AMIS_percent %>% select(-MFs), BMIS_percent %>% select(-MFs)) %>%
  mutate(Originality_Index = PercentofBMIS/PercentofAMIS)
  
Summarylong <- left_join(AMIS_percent %>% select(-MFs), BMIS_percent %>%
                           select(-MFs))%>% 
  gather(key = "type", value = "Percent", -FinalBMIS)

g <- ggplot() +
  geom_col(data = Summarylong , aes(x = type, y = Percent))+
  facet_wrap(~FinalBMIS)

g

# write.csv(Summary, "AMISvsBMIS.csv")

# ggsave("AMISvsBMIS.pdf", plot = g, device = "pdf", width = 10, height = 10, units = "in")

```


###Get all the data back - and keep only the MF-MIS match set for the BMIS

```{r}
#Add a column to the longdat that has important information from the FullDat_fixed, then only return data that is normalized via B-MIS normalization
BMIS_normalizedData <- FullDat_fixed %>%
  filter(MIS == FinalBMIS) %>%
  select(MassFeature, Orig_RSD, del_RSD, percentChange, FinalBMIS, FinalRSD) %>%
  right_join(longdat) %>%
  filter(MIS == FinalBMIS) %>%
  left_join(., dat %>% select(Run.Cmpd, runDate, type, SampID, replicate))

new.filename <- paste("BMISd",filename,sep="_")
write.csv(BMIS_normalizedData, new.filename)
print(paste("Your BMIS results are saved as", new.filename))
print("The output of the B-MIS normalized data is in the column named PooPlusModel.")
```


#Senitivity analysis - Do we have enough internal standards?  
You need to have run BMIS and have rsd.total from the BMIS results
And you need the IS.list
This might take a bit...several minutes on a medium-high powered computer
```{r Do we have enough Internal Standards?, error=FALSE, message=FALSE, fig.width=8, fig.height=8}

IS.list <- unique(IS.dat$MassFeature)

FullDat_fixedCulled <- FullDat_fixed %>%
  filter(FinalRSD < 0.2)

rsd.sens <- FullDat_fixedCulled %>% 
  select(`MassFeature`:accept_MIS) %>% 
  select(-Poo.Picked.IS) %>%
  filter(!`MassFeature` %in% IS.list) %>% #Gets rid of IS as mass features
  mutate(ISused = MIS %>% str_replace(".Norm.Area", ""))

IS_number <- NA
BMIS_failed <- as.data.frame(IS_number)
BMIS_failed$Iteration <- NA
BMIS_failed$Percent_Fail <- NA
BMIS_failed$RSD <- NA
BMIS_failed_empty <- BMIS_failed

for (i in 1:length(IS.list)){
ISsublist <- combn(IS.list, i, simplify = FALSE)
print(paste( "On iteration ",i, " of ", length(IS.list)))

#Turn into a loop for length of ISsublist, every combination of IS at each #
if (length(ISsublist) > 50) {ISsublist <- sample(ISsublist, 50)}
for (j in 1:length(ISsublist)){
ISsubs <- c(ISsublist[[j]], "Inj_vol")
subrsd <- rsd.sens %>%
  filter(ISused %in% ISsubs) 
subrsd <- subrsd %>% left_join(subrsd %>%
                                 group_by(MassFeature) %>%
                                 summarise(Poo.Picked.IS = unique(MIS)[which.min(RSD_ofPoo)][1]))


#Change the BMIS to "Inj_vol.Norm.Area" if the BMIS is not an acceptable - Adds a column that has the BMIS, not just Poo.picked.IS.  Then newpoodat_BMIS is the final results
NoAMIS_newpoodat_BMIS <- subrsd %>%
  filter(MIS == Poo.Picked.IS) %>% #Pulling out just the pooplus is matches
  filter(accept_MIS == "FALSE") %>%
  mutate(FinalBMIS = "Inj_vol.Norm.Area",
         FinalRSD = Orig_RSD)

newpoodat_BMIS <- subrsd %>%
              filter(MIS == Poo.Picked.IS) %>% 
              filter(accept_MIS == "TRUE") %>%
              mutate(FinalBMIS = Poo.Picked.IS,
                     FinalRSD = RSD_ofPoo) %>%
  bind_rows(NoAMIS_newpoodat_BMIS)

BMIS_failed_iteration <- BMIS_failed_empty
BMIS_failed_iteration$IS_number[1] <- i
BMIS_failed_iteration$Iteration[1] <- j

BMISFail <- newpoodat_BMIS %>% filter(FinalBMIS == "Inj_vol.Norm.Area")
BMIS_failed_iteration$Percent_Fail[1] <- length(BMISFail$FinalBMIS)/length(newpoodat_BMIS$Poo.Picked.IS)

BMIS_failed_iteration$RSD[1] <- median(newpoodat_BMIS$FinalRSD)###Add final RSD
BMIS_failed <- rbind(BMIS_failed, BMIS_failed_iteration)

}

}
BMIS_sensResults <- list()
BMIS_sensResults[[1]] <- BMIS_failed


BMIS_failed_summary <- BMIS_failed %>%
  group_by(IS_number) %>%
  summarise(MedPercent = median(Percent_Fail),
            MedRSD = median(RSD))

BMIS_sensResults[[2]] <- ggplot(data = BMIS_failed, aes(x = IS_number, y = RSD)) +
  geom_point()+
  geom_point(data = BMIS_failed_summary, aes(x = IS_number, y = MedRSD), size = 5)

BMIS_sensResults[[3]] <- ggplot(data = BMIS_failed, aes(x = IS_number, y = Percent_Fail)) +
  geom_point() +
  geom_point(data = BMIS_failed_summary, aes(x = IS_number, y = MedPercent), size = 5)

BMIS_failed_summary
BMIS_sensResults[[2]]
BMIS_sensResults[[3]]

save(BMIS_sensResults, file = "BMIS_SensResults.RData")


#Get % of MFs that have less than 10% RSD (and will never pick a feature)
Try <- FullDat_fixedCulled %>%
  select(MassFeature, Orig_RSD) %>%
  unique() %>%
  filter(Orig_RSD < 0.1)

```



#Plot to look at which BMIS was chosen
Requires unsplit.data.join
```{r ,  error = FALSE, warning=FALSE, message=FALSE, fig.width=8, fig.height=6}
xcms.subset <- xcms.dat %>% select(mz, rt, MassFeature)
IS.dat <- ISdatfull %>%
  mutate(rt = `Retention Time`*60, 
         mz = `Precursor Mz`,
         MassFeature = `Precursor Ion Name`)



IStoplot <- IS.dat %>%
  select(MassFeature, rt, mz) %>%
   group_by(MassFeature) %>%
  summarise(rt = mean(rt), mz = mean(mz)) %>%
   mutate(type = "IS",
         PooPlusModel.IS = MassFeature)



dattoplot <- unsplit.dat.join %>%
  select(MassFeature, PooPlusModel.IS) %>%
  filter(!is.na(PooPlusModel.IS)) %>%
  mutate(PooPlusModel.IS = PooPlusModel.IS %>% str_replace(".Norm.Area", "")) %>%
  left_join(xcms.subset) %>%
  filter(!is.na(mz)) %>%
  unique() %>%
  mutate(type = "analyte") %>%
  rbind(IStoplot)

NoBMIStoplot <-dattoplot %>%
  filter(PooPlusModel.IS == "Inj_vol")

dattoplot <- dattoplot %>%
  filter(!PooPlusModel.IS == "Inj_vol")

ggplot(dat = dattoplot, aes (x = rt, y = mz, text = MassFeature))+
  geom_point(aes(fill = PooPlusModel.IS, color = type, alpha = type),
                 pch=21, size = 2, alpha = 0.6) +
  scale_colour_manual(values=c("white", "black"))+ 
  annotate("text", x = mean(range(dattoplot$`Retention Time`)),  y = max(dattoplot$Parent),  label= "Black dots are where no BMIS were selected")+
  geom_point(dat = NoBMIStoplot, aes(x = `rt`, y = mz ), 
             colour = "black", size = 2) +
  theme(legend.text=element_text(size=8))

```

Black dots are where no BMIS were selected
